---
title: Equity strategy overview
format: html
editor: source

execute:
  echo: true
---

The Tampa Bay Estuary Program (TBEP) is developing a methodology as part of its new Equity Strategy to identify historically underserved (i.e., disadvantaged) communities across the Tampa Bay watershed to support the goals of the White House's [Justice40 Initiative](https://www.whitehouse.gov/environmentaljustice/justice40/){target="_blank"} (EO 14008) and the Environmental Protection Agency's [Equity Action Plan](https://www.epa.gov/environmentaljustice/equity-action-plan){target="_blank"} (EO 13985). The purpose of this initiative is to ensure TBEP delivers equitable and fair access to the benefits from environmental programs for all communities.

We have created this webpage to increase transparency and reproducibility of the approach guiding our Equity Strategy. We have provided instructions for downloading, cleaning, and analyzing the data that will be used to identify underserved and overburdened communities in Tampa Bay. The methods described on the subsequent pages are presented in R coding language, but users who prefer working with GIS software (e.g., ArcGIS or QGIS) should also be able to reproduce the maps by following along in the descriptions of each step.

## What do we consider?

To identify underserved communities in Tampa Bay, we are guided by recommendations provided by the Environmental Protection Agency (EPA), as well as our own local knowledge and experience with the diverse population of more than 3 million people across the Tampa Bay watershed.

The 2021 [Bipartisan Infrastructure Law (BIL)](%22https://www.govinfo.gov/content/pkg/PLAW-117publ58/pdf/PLAW-117publ58.pdf%22), as implemented by EPA through NEPs, provides approximately \$909,800 in annual funding to each NEP over 2022-2026. EPA has a goal to ensure that at least 40% of the benefits from BIL-funded projects during this period will flow to underserved (or disadvantaged) communities across all NEPs.

The EPA uses the EJScreen Supplemental Demographic Index to identify disadvantaged communities. The Index is based on an average of the following five demographic variables:

-   Percent classified as low income
-   Percent unemployed
-   Percent linguistically isolated (i.e., limited English speaking)
-   Percent with less than a high school education
-   Percent with low life expectancy

TBEP applies a more inclusive definition of underserved communities to identify opportunities for increasing environmental justice throughout the watershed. For this broader definition, we define "underserved" communities as those that fall within the 80th percentile (or higher) nationally for two or more of the five demographic variables listed above.

Additionally, our Equity Strategy recognizes the importance of tailoring TBEP's environmental programs to address the most significant challenges facing different underserved communities in Tampa Bay. In an effort to promote environmental justice within our region, our approach also characterizes underserved communities according to the unique burdens they are facing that could realistically be mitigated by TBEP activities. We define "overburdened" communities as those that fall within the 80th percentile (or higher) nationally for one or more of the following 12 environmental justice issues:

-   *Climate change*: projected agricultural loss, projected flood risk
-   *Nature deprivation*: lack of green space
-   *Air pollution*: PM2.5, traffic proximity/volume
-   *Water pollution*: wastewater discharge
-   *Other pollution*: hazardous waste facilities, underground storage tank release, Superfund sites, brownfield sites, phosphate mines
-   *Legacy effects*: historic underinvestment/redlining

## Contact us

This methodology may be adapted by other National Estuary Programs or local government partners developing their own equity strategies. Along with our descriptions of the methodology, we also briefly highlight the rationale behind our decisions, which may or may not be relevant or appropriate to other organizations. If you have questions or other feedback regarding the methodology, contact Dr. Blake Simmons at [bsimmons\@tbep.org](mailto:bsimmons@tbep.org).

## Underserved communities in Tampa Bay

```{r}
#| echo: false
library(sf)
library(mapview)
library(RColorBrewer)
library(dplyr)

load(file = 'data/underserved_tract.RData')
load(file = 'data/underserved_dbasins.RData')
load(file = 'data/tb_equity.RData')

underserved_tract %>%
  filter(underserved == "Yes") %>%
  mapview(col.regions = "red", layer.name = "Underserved Census Tracts")
```

## Drainage basins containing underserved communities

```{r}
#| echo: false

mapviewOptions("basemaps.color.shuffle" = FALSE)

underserved_dbasins %>%
  filter(pct_under > 0) %>%
  mapview(zcol = "pct_under", col.regions = brewer.pal(6, "Reds"), layer.name = "Underserved Tracts (% of DB)", label = "WATERBODY_NAME")
```

## Underserved and overburdened communities in Tampa Bay

```{r}
#| echo: false

mapviewOptions("basemaps.color.shuffle" = FALSE)

tb_equity %>%
  mapview(zcol = "thresholdEJ_N", col.regions = brewer.pal(8, "YlOrRd"), layer.name = "No. of Burdens", label = "EJissues1")
```

Navigate through the Methods pages to create this data and replicate these maps.
