---
title: "Downloading and cropping EJScreen data"
format: html
editor: source
---

Load the required R packages (install first as needed, tbeptools installation instructions are [here](https://tbep-tech.github.io/tbeptools/))
.
```{r}
library(tbeptools)
library(sf)
library(mapview)
library(dplyr)
library(RColorBrewer)
```

Download the relevant file from EJScreen.  All files available at <https://gaftp.epa.gov/EJSCREEN/2022/?C=S;O=A>.  The file is downloaded to a temporary directory. 

```{r}
# url with zip gdb to download
urlin <- 'https://gaftp.epa.gov/EJSCREEN/2022/EJSCREEN_2022_Supplemental_with_AS_CNMI_GU_VI_Tracts.gdb.zip'

# download file
tmp1 <- tempfile(fileext = "zip")
download.file(url = urlin, destfile = tmp1)
```

Unzip the geodatabase that was downloaded to a second temporary directory. 

```{r}
# unzip file
tmp2 <- tempfile()
utils::unzip(tmp1, exdir = tmp2, overwrite = TRUE)
```

Read the polygon layer from the geodatabase.

```{r}
# get the layers from the gdb
gdbpth <- list.files(tmp2, pattern = '\\.gdb$', full.names = T)
lyrs <- st_layers(gdbpth)$name

# read the layer
dat <- st_read(dsn = gdbpth, lyrs)
```

Intersect the layer with the Tampa Bay watershed.

```{r}
# intersect the layer with the tb watershed
dattb <- dat %>% 
  st_transform(crs = st_crs(tbshed)) %>% 
  st_make_valid() %>% 
  st_intersection(tbshed)
```

View the data using mapview.

```{r}
#| out-width: "100%"
# view the layer
mapview(dattb)
```
```{r}
dattbcopy <- dattb %>%
  filter(ACSTOTPOP > 0) %>%
  mutate(threshold_income = ifelse(P_LWINCPCT >= 80, 1, 0),
         threshold_unempl = ifelse(P_UNEMPPCT >= 80, 1, 0),
         threshold_lingui = ifelse(P_LNGISPCT >= 80, 1, 0),
         threshold_educat = ifelse(P_LESHSPCT >= 80, 1, 0),
         threshold_lifexp = ifelse(P_LIFEEXPCT >= 80, 1, 0)) %>%
  rowwise() %>%
  mutate(threshold_N = sum(threshold_income,threshold_unempl,threshold_lingui,threshold_educat,threshold_lifexp, na.rm = TRUE)) %>%
  mutate(underserved = ifelse(threshold_N > 1, "Yes", "No"))
mapview(dattbcopy, zcol = "threshold_N")


mapview(dattbcopy, zcol = "underserved")

```

The layer can be saved as an RData object if needed.  Size is very small (~1mb).

```{r}
#| eval: false
# save the layer as an RData object (~1mb)
save(dattb, file = 'data/dattb.RData')
```

Unlink the temporary files so they're deleted when you're done. 

```{r}
# remove temp files
unlink(tmp1, recursive = TRUE)
unlink(tmp2, recursive = TRUE)
```

